Fail Faster

### 2025-03-21 paper/paper.tex
done through 227

While many DSL designs are possible, the standard design for QuickCheck-style
generators is via an embedded {\em monadic} language. For the purposes of this
paper, we use the syntax and types from OCaml's \bq library, which is presented in
Figure~\ref{fig:bq-api}, but these kinds of DSLs can also be found in languages
like Haskell, Scala, Python, and many more.
The library provides some basic generators, for
example \texttt{gen\_int} for generating random integers in a range, along with
the ``monadic interface'' consisting of \texttt{return} and \texttt{bind}.  The
generator \texttt{return x} is the constant generator, always generating the
value \texttt{x}. Running a generator \texttt{bind g k} runs the generator
\texttt{g}, producing a value \texttt{a}, and then runs the generator \texttt{k
a}.
\iflater
\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also
think we should completely omit named arguments in the paper so we don't confuse
with splices.}
\fi

Together, these three functions are the bare minimum for constructing arbitrary
random data generators: \texttt{gen\_int} provides a base source of randomness,
and \texttt{return} and \texttt{bind} allow generators to be composed to create
larger, more complex generators. Figure~\ref{fig:simple-bind} shows a generator
built with these operations; it first samples an int between $0$ and $100$,
names it $x$, samples another between $0$ an $x$, and then returns the pair of
them both.

\begin{figure}[h]
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  Bq.bind (Bq.gen_int 0 100) (fun x ->
    Bq.bind (Bq.gen_int 0 x) (fun y ->
      Bq.return (x,y)))
\end{lstlisting}
\caption{A simple generator using \texttt{bind} explicitly.}\label{fig:simple-bind}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = Bq.gen_int 0 100 in
  let%bind y = Bq.gen_int 0 x in
  return (x,y)
\end{lstlisting}
\caption{An equivalent generator using macros for \texttt{bind}.}\label{fig:simple-macro}
  \end{subfigure}
  \caption{Simple monadic generators for a pair of ordered integers.}\label{fig:simple-pair}
\end{figure}

Most languages in which monadic APIs are common expose some sort of syntactic
sugar for them. In OCaml\footnote{OCaml actually has a few ways to implement
monadic syntax; this is the one provided by Jane Street's libraries.}, this
looks like \texttt{let\%bind x = e in e'}, which desugars to
\texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
generator as before, now written the monadic syntax.

@@

Generator libraries also include other functions that make generator
construction easier; some examples of these are also shown in
Figure~\ref{fig:bq-api}.  The \texttt{weighted\_union} function is a
particularly well-used one: it makes making a weighted choice between different
generators, allowing the developer to combine different sub-generators into a
single program and tune the data distribution. Also important are functions like
\texttt{size} and \texttt{with\_size} that are used to control the sizes of
generated values and \texttt{fixed\_point} that is used to define recursive
generators.

@@

The generator in Figure~\ref{fig:tree_of} uses all of the aforementioned
features. It uses \texttt{fixed\_point} to define a recursive generator that
reads the current value of \texttt{size} to determine how to generate a tree.
The generator uses \texttt{weighted\_union} to make a random choice between an
empty tree and a node, choosing a node with weight proportional to the current
size and choosing a leaf otherwise. When generating a node, the generator uses
\texttt{with\_size} to reduce the value of the size parameter for future
iterations.

\subsection{Abstraction Overhead of Generator DSLs}
Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
for an instance of the reader monad \cn{}. 
% \bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.

Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
\bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.

@@

Just how much run-time overhead does this monadic abstraction introduce? To
illustrate, let's return to our running example of a constrained pair of
integers, written in both \bq{} and ScalaCheck.
Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
The second versions (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
are semantically identical to the first, but have been rewritten
by (1) inlining all generator combinator definitions, and then (2)
repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} --- where an anonymous function is defined and then immediatly called~\footnote{This is usually known as a ``$\beta$-redex''.
Note that when reducing, we do not reduce to \lstinline{e[e'/x]}, as this does not preserve the order of effects.
}
--- to \lstinline{let x = e' in e}.

\begin{figure}
\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = (Bq.gen_int 0 100) in
  let%bind y = (Bq.gen_int 0 x) in
  Bq.return (x,y)

let int_pair_inlined : int -> SR.t -> int * int =
  fun _ sr ->
    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
    let y = Splittable_random.int sr ~lo:0 ~hi:x in
    (x,y)

def intPair : Gen[(Long,Long)] = for {
  x <- Gen.choose(0,1000)
  y <- Gen.choose(0,x)
} yield (x,y)

def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
}
\end{lstlisting}
\caption{Int Pair Generators in \bq{} and ScalaCheck}
\label{fig:overhead-explanation-code}
\end{figure}

The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes (on average) half as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
show an even more dramatic performance boost for inlining in this manner.

Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler.
Indeed, compilers of effectful and strict functional languages (including JIT
compilers, in the case of Scala 3)
use heuristics to determine if and when to perform this particular kind of
simplification~\footnote{As we discuss in Section~\ref{section:other-langs}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}.  Even in cases as simple as
Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.

>>>>

The standard design for QuickCheck-style
generators is via an embedded {\em monadic} language. We use the syntax and types from OCaml's \bq library, which is presented in
Figure~\ref{fig:bq-api}, but similar DSLs can also be found in languages
like Haskell, Scala, Python, and many more.
The library provides some basic generators, for
example \texttt{gen\_int} for generating random integers in a range, along with
the functions \texttt{return} and \texttt{bind}.  The
generator \texttt{return x} is the constant generator, always generating the
value \texttt{x}. Running a generator \texttt{bind g k} runs the generator
\texttt{g}, producing a value \texttt{a}, and then runs the generator \texttt{k
a}.
\iflater
\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also
think we should completely omit named arguments in the paper so we don't confuse
with splices.}
\fi

Together, these three functions are the bare minimum for constructing arbitrary
random data generators: \texttt{gen\_int} provides a base source of randomness,
and \texttt{return} and \texttt{bind} allow generators to be composed to create
larger, more complex generators. Figure~\ref{fig:simple-bind} shows a generator
built with these operations; it first samples an int between $0$ and $100$,
names it $x$, samples another between $0$ and $x$, and then returns the pair of
them.

\begin{figure}[h]
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  Bq.bind (Bq.gen_int 0 100) (fun x ->
    Bq.bind (Bq.gen_int 0 x) (fun y ->
      Bq.return (x,y)))
\end{lstlisting}
\caption{A simple generator using \texttt{bind} explicitly.}\label{fig:simple-bind}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = Bq.gen_int 0 100 in
  let%bind y = Bq.gen_int 0 x in
  return (x,y)
\end{lstlisting}
\caption{An equivalent generator using the macro for \texttt{bind}.}\label{fig:simple-macro}
  \end{subfigure}
  \caption{Simple monadic generators for a pair of ordered integers.}\label{fig:simple-pair}
\end{figure}

Most languages in which monadic APIs are common expose some sort of syntactic
sugar for them. In OCaml\footnote{OCaml actually has a few ways to implement
monadic syntax; this is the one provided by Jane Street's libraries.}, this
looks like \texttt{let\%bind x = e in e'}, which desugars to
\texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
generator, written the monadic syntax.

@@

Generator libraries also include other functions that make generator
construction easier; some examples of these are included in
Figure~\ref{fig:bq-api}.  The \texttt{weighted\_union} function is a
particularly well used one: it makes making a weighted choice between different
generators, allowing the developer to combine different sub-generators into a
single program and tune the data distribution. Also important are functions like
\texttt{size} and \texttt{with\_size} that are used to control the sizes of
generated values and \texttt{fixed\_point} that is used to define recursive
generators.

@@

The generator in Figure~\ref{fig:tree_of} uses all of these
features. It uses \texttt{fixed\_point} to define a recursive generator that
reads the current value of \texttt{size} to determine how to generate a tree.
It uses \texttt{weighted\_union} to make a random choice between an
empty tree and a node, choosing a node with weight proportional to the current
size and choosing a leaf otherwise. When generating a node, it uses
\texttt{with\_size} to reduce the value of the size parameter for future
iterations.

\subsection{Abstraction Overhead of Generator DSLs}
Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
\jwc{introduce what SR is}
taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
and returning a generated value \texttt{'a}. It is an invariant of the generator library that every function of this type is
deterministic: for a fixed size and seed, it will always return the same value, so all of the randomness in testing comes from varying the initial seed.
The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
for an instance of the reader monad \cn{} \jwc{explain better}.
% \bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here. \jwc{lift this explanation, explain what splittable random is before here.}

Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
\bcp{Sentence starts with lowerase} \bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
\bcp{This discussion should go elsewhere...}

@@

Just how much run-time overhead does this monadic abstraction introduce? To
illustrate, let's return to our running example of a constrained pair of
integers, written in both \bq{} and ScalaCheck.
Figure~\ref{fig:simplify} shows two versions of the generator, written in both languages.
The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
The second versions (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
are semantically identical to the first, but have been rewritten
by (1) inlining all generator combinator definitions, and then (2)
repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} where an anonymous function is defined and then immediately called, to \lstinline{let x = e' in e}.

\begin{figure}[h]
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = (Bq.gen_int 0 100) in
  let%bind y = (Bq.gen_int 0 x) in
  Bq.return (x,y)

let int_pair_inlined : int -> SR.t -> int * int =
  fun _ sr ->
    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
    let y = Splittable_random.int sr ~lo:0 ~hi:x in
    (x,y)
\end{lstlisting}
\caption{Simplifying a Generator in OCaml}\label{fig:simplify-ocaml}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
\begin{lstlisting}
def intPair : Gen[(Long,Long)] = for {
  x <- Gen.choose(0,1000)
  y <- Gen.choose(0,x)
} yield (x,y)

def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
}
\end{lstlisting}
\caption{Simplifying a Generator in Scala}\label{fig:simplify-scala}
  \end{subfigure}
  \caption{Simplifying Generators}\label{fig:simplify}
\end{figure}

% \begin{figure}
% \begin{lstlisting}
% \end{lstlisting}
% \caption{Int Pair Generators in \bq{} and ScalaCheck}
% \label{fig:overhead-explanation-code}
% \end{figure}

The performance impact of this simplification is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes on average half as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
show an even more dramatic performance boost.

Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler.
Indeed, compilers of effectful and strict functional languages (including JIT
compilers, in the case of Scala 3)
use heuristics to determine if and when to perform this particular kind of
simplification~\footnote{As we discuss in Section~\ref{section:other-langs}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}.  Even in cases as simple as
Figure~\ref{fig:simplify}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.
